---
title: "Written Report"
author: "Outliers: Matthew Hawkins, Aston Yong, Catherine Ning"
date: "November 11, 2021"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Introduction and Data

For this project, we are looking into the Himalayan database, which is a
compilation of all expeditions that have climbed in the Nepal Himalaya.
The data cover all expeditions from 1905 through Spring 2019 to more
than 465 significant peaks in Nepal. Also included are expeditions to
both sides of border peaks such as Everest, Cho Oyu, Makalu and
Kangchenjunga as well as to some smaller border peaks. Data on
expeditions to trekking peaks are included for early attempts, first
ascents and major accidents. The main research question we are hoping to
solve involves looking into various different factors which can predict
the safety and success of the expedition.

Many active climbers view the Himalayas as the proverbial "summit" of
their climbing experience. These mountains provide a demanding and often
dangerous task so, the most important goal for climbers is safety and
success. We want to create a model that find the most significant
factors for safe and successful expeditions so that future climbers can
predict probabilities of safe, successful climbs. Climbing the Himalayas
is still a dangerous and risky activity, and increased congestion is
changing the dynamic of climbing
(<http://graphics.reuters.com/NEPAL-EVEREST/0100B4S22JR/index.html>).
Therefore, examining these dataset predictors and drawing insights from
them will hopefully improve mountain climbing preparedness.

We wish to explore the following research question:

What expedition factors are most predictive of safe and successful
expeditions and what is the probability of success given these
predictors?

We hypothesize expeditions with conditions like lower peaks,
historically climbed peaks, more members, warm seasons, younger average
age, and more recent year to have greater probability of safe success.

The data cover all expeditions from 1905 through Spring 2019 to more
than 465 significant peaks in Nepal. It is divided into 3 tables, one
for peaks, expeditions, and members respectively. The peaks.csv file
contains information about each Himalayan mountain with data on peak
name, whether it has been climbed, height, and other variables. The
expeditions.csv file contains data on individual expeditions, with
variables like member count, season, and number of deaths. The
members.csv file is the most narrow-scoped table with demographic and
personal information for each member on an expedition. Multiple members
can participate in one expedition, and multiple expeditions can climb
one peak.

The database is based on the expedition archives of Elizabeth Hawley, a
longtime journalist based in Kathmandu, and it is supplemented by
information gathered from books, alpine journals and correspondence with
Himalayan climbers. The database is updated bi-annuallly, and member
information comes from submitted permit applications.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages}
library(tidyverse)
library(broom)
library(patchwork)
library(knitr)
library(ggfortify) 
library(yardstick)

# load other packages as needed
```

```{r load-data}
# load all data sets 
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv')
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
peaks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/peaks.csv')
```

```{r, include=FALSE}
members %>% filter(!is.na(age)) %>% group_by(expedition_id) %>% summarize(mean_age = mean(age), member_count = n() ) %>% kable(digits = 3)
```

```{r}
expeditions <- expeditions %>%
  mutate(success_expedition = factor(if_else((termination_reason == "Success (main peak)") | (termination_reason == "Success (subpeak)"), "1", "0")))
```

We created our response variable,
sucess_expedition, defining an expedition as a success or a failure
based on the reason for the expedition's termination (Success (main peak), Success
(subpeak), accident, bad weather, etc). A successful expedition has the value "1" 
and an unsuccessful expedition has the value "0".


```{r fig.width = 4, fig.height = 4}
ggplot(data = expeditions, mapping = aes(x = success_expedition)) +
  geom_bar() + 
  labs(title = "Distribution of Expedition Outcomes", 
       x = "",
       y = "Number of Expeditions") +
  scale_x_discrete(labels = c("1" = "Success", "0" = "Failure"))
```

```{r}
expeditions %>%
  count(success_expedition) %>%
  kable(digits = 3)
```

```{r}
success_prop <- 5707 / 10364
failure_prop <- 4657 / 10364
```

Based on the above count and calculation, 55.1% of the 10364 expeditions
were classified as successful. On the other hand, 44.9 of the 10364
expeditions were classified as unsuccessful.

```{r load-data—1}
# load data set
members_grouped <- read_csv("data/members_grouped.csv")
```

First of all, we join grouped members info and peak height into
expeditions.

```{r }
merged_data_1<-merge(x=members_grouped,y=expeditions,by="expedition_id",all.x=TRUE)
merged_data_2<-merge(x=merged_data_1,y=peaks,by="peak_id",all.x=TRUE)
```

We selected all the possible predictor variables we are planning to use,
including peak height, member count, number hired members, oxygen use,
year, season, agency, member deaths, and hired deaths. We removed all
NAs from trekking_agency. From the members.csv file, average member age
on an expedition calculated from grouping by expedition id.

```{r include=FALSE}
merged_data <- merged_data_2 %>%
  select(height_metres, member_count, hired_staff, oxygen_used, year, season,
       trekking_agency, member_deaths, hired_staff_deaths, mean_age, success_expedition)
merged_data <- merged_data %>%
  filter(!is.na(trekking_agency))
merged_data
```

```{r}
#length(unique(merged_data$trekking_agency))
```

There are over 800 trekking agencies, so to simplify the data, we only
kept the top 10 agencies and set trekking_agency to Other for any other
agency. To make the model intercept interpretation meaningful, we also
mean centered the continuous variables for mean age, mountain height,
and member count.

```{r}
merged_data <- merged_data %>%
  mutate(trekking_agency = fct_lump(trekking_agency, 10, other_level = "Other",
  ties.method = "first"))
```

```{r, include=FALSE}
merged_data %>% count(trekking_agency) %>% arrange(desc(n))
merged_data %>% count(year) %>% arrange(desc(year))
merged_data %>% count(oxygen_used)
merged_data %>% count(height_metres)
```

```{r fig.width = 4, fig.height = 4}
ggplot(data = expeditions, mapping = aes(x = oxygen_used)) +
  geom_bar() + 
  labs(title = "Oxygen Use Distribution", 
       x = "",
       y = "Number of Expeditions") +
  scale_x_discrete(labels = c("TRUE" = "Used", "FALSE" = "Not Used"))
```

In the above bar graph, we can see the portion of expeditions in which
oxygen was used, which we believe to be a potentially important
predictor variable. As the graph shows, a little less than a third of
expeditions used oxygen.

```{r fig.width = 4, fig.height = 4}
ggplot(data = expeditions, mapping = aes(x = members)) + 
  geom_histogram() +
  labs(title = "Number of Members Histogram", x = "Number of Members in Each Group")
```

```{r}
expeditions %>%
  summarise(mean = mean(members), sd = sd(members), IQR = IQR(members)) %>%
  kable(digits = 3)
```

The above histogram shows the distribution of the number of members in
each group, another variable which we believe may be a significant
predictor of the odds of success of an expedition. The distribution is
right-skewed and seems like it may have a few outlier expeditions that
have greater than 25 members in the group. The distribution is centered
with a mean of 5.953 members, and it has a spread characterized by a
standard deviation of 5.428 and an interquartile range of 6.


```{r, include = FALSE}
ggplot(data = merged_data, aes(x = height_metres)) +
geom_histogram(binwidth = 50) +
labs(title = "Height Distribution")
```

We also mapped out the distribution of season. We can notice that for spring and autumn, the numbers of expeditions are over 4,000 much higher than those of summer and winter, below 200. 

```{r fig.width = 4, fig.height = 4}
ggplot(data = merged_data, mapping = aes(x = season, fill = season)) +
  geom_bar() + 
  labs(title = "Season Distribution", 
       x = "",
       y = "Number of Expeditions") +
  scale_x_discrete(labels = c("Autumn", "Spring", "Summer", "Winter"))
```

```{r}
merged_data %>%
  count(trekking_agency) %>%
  kable(digits = 3)
```

According to the graph, the "other" agency has the most trekking agencies, about
5500, while among the following trekking agencies, Thamserku and Asian trekking 
have relatively higher number of trekking agencies compared to the rest of the agencies.

```{r fig.width = 12, fig.height = 4}
ggplot(data = merged_data, mapping = aes(x = trekking_agency, fill = trekking_agency)) +
  geom_bar() + 
  labs(title = "Agency Distribution", 
       x = "",
       y = "Number of Expeditions")+
  coord_flip()
```

## 2. Methodology

Because we want to predict success, a categorical response, we use
logistic regression to model the log odds of expedition success versus
failure. Probability of success must be between 0 and 1. We used forward
selection and BIC to build our model and determine signficant predictor
variables of odds of success. Forward selection model using BIC produces
the following model:

```{r}
int_only <-glm(success_expedition ~ 1, data = merged_data, family = binomial)
#tidy(int_only) %>% kable(digits = 3)
full_model <-glm(success_expedition ~ ., data = merged_data,family = binomial)
#tidy(full_model)  %>% kable(digits = 3)
```

```{r, include=FALSE}
forward_select_bic <- step(int_only, scope = formula(full_model), direction = "forward", k = log(nrow(merged_data)))
```

```{r}
tidy(forward_select_bic) %>% kable(digits = 5)
```

```{r}
merged_data <- merged_data %>%
  mutate(year = year-1971,
         height_metres_cent = height_metres-mean(height_metres),
         member_count_cent = member_count-mean(member_count),
         mean_age_cent = mean_age-mean(mean_age),
         oxygen_used = factor(oxygen_used),
         season = factor(season),
         trekking_agency = factor(trekking_agency)
         )
```

```{r}
library(Stat2Data)
```

In the predictor variables' empirical logit plots (Appendix 5.1), we observe 
interaction effects between oxygen use and the quantitative predictors peak 
height and year. We therefore add interaction terms into our model. There is also
a clear quadratic relationship between log odds of success and mean_age_cent, so
we transformed mean_age_cent into a quadratic term. This is expected as
too young or too old climbers will be less prepared or fit. Year also
showed possible quadratic relationships, but squaring year did not
change the empirical logit plot. We decided to keep year unchanged. The
predictor year in our model with interaction terms and quadratic
transformed mean age had a p-value greater than 0.05, so we performed a
drop in deviance test. The test p-value is near zero, so we keep year in
the model.

```{r}
merged_data <- merged_data %>% mutate(
  mean_age_cent_sq = mean_age_cent^2 #,
#  year_sq = year^2
)
forward_select_bic <- glm(success_expedition ~ oxygen_used + height_metres_cent + member_count_cent +
                            mean_age_cent + mean_age_cent_sq + hired_staff_deaths + year + 
                            height_metres_cent*oxygen_used + year*oxygen_used, data = merged_data,
                          family = binomial)
```

```{r results = "hide"}
reduced_model <-glm(success_expedition ~ oxygen_used + height_metres_cent + member_count_cent +
                            mean_age_cent + mean_age_cent_sq + hired_staff_deaths + 
                            height_metres_cent*oxygen_used, data = merged_data,
                          family = binomial)
```


```{r}
tidy(forward_select_bic, conf.int = TRUE) %>% kable(digits = 5)
```

```{r}
anova(reduced_model, forward_select_bic, test = "Chisq") %>%
  kable(digits = 3)
```

```{r}
par(mfrow = c(1,2))
emplogitplot1(success_expedition ~ mean_age_cent_sq, data = merged_data, ngroups = 8) 
emplogitplot2(success_expedition ~ mean_age_cent_sq + oxygen_used, data = merged_data, 
              ngroups = 8, putlegend = "bottomright")

```

Checking model conditions:

1.  Linearity: The empirical logit plots (above and Appendix 5.1) show a general linear
    relationship between the log odds of success and year, peak height,
    and hired staff deaths. There is a strong linear relationship
    between log odds of success and member count and mean age squared.
2.  Randomness: randomness is satisfied because the data is just a
    record of all Himalayan expeditions, so the data collection is
    representative of the population and is not a particular subset.
3.  Independence: independence condition is satisfied because
    expeditions are separate and odds of success for one expedition do
    not tell anything about odds of success for another expedition.

Multicolinearity is not an issue, as the Variance Inflation Factor (VIF)
is greater than 10 only for oxygen_used and its interaction with year.
This is expected as an interaction term needs the associated main
effects.

```{r}
library(rms)
vif(forward_select_bic) %>% kable(digits = 5)
```

## 3. Results

Our final model is displayed here:

```{r}
tidy(forward_select_bic, conf.int = TRUE) %>% kable(digits = 5)
```

```{r}
forward_select_bic_aug <- augment(forward_select_bic, type.predict = "response")


# Need to put 1 as the first level
forward_select_bic_aug <- forward_select_bic_aug %>%
  mutate(success_expedition = fct_relevel(success_expedition, c("1", "0")))
# calculate sensitivity and specificity at each threshold
roc_curve_data <- forward_select_bic_aug %>%
  roc_curve(success_expedition, .fitted) 
# plot roc curve
autoplot(roc_curve_data)

forward_select_bic_aug %>%
  roc_auc(success_expedition, .fitted) %>%
  pull(.estimate)
```

**Interpretations:**

The key predictors of expedition success are peak height, member count, mean 
member age, staff deaths, year, and oxygen use.

Intercept: For an expedition in 1971 with mean peak height of
`r round(mean(merged_data$height_metres),3)` meters,
`r round(mean(merged_data$member_count),3)` group members, mean age of
`r round(mean(merged_data$mean_age),3)`, 0 staff deaths, and no oxygen
use, we would expect the odds of a successful expedition to be
`r round(exp(-0.12853), 3)`.

Height_metres_cent: For each additional meter in the height of the peak,
we would expect, on average, that the odds of a successful expedition
are multiplied by `r round(exp(-0.00063), 3)`, holding all else
constant.

Member_count_cent: For each additional member in the group, we would
expect, on average, that the odds of a successful expedition are
multiplied by `r round(exp(0.06468), 3)`, holding all else constant.

Mean_age_cent: Because of the quadratic term, mean_age_cent must be
interpreted using the overall effect. The effect of mean_age_cent on log
odds equals zero at mean centered mean_age of -5.285, which equals an expedition
mean age of `r round(mean(merged_data$mean_age),3) - 5.285`. The negative coefficient
of the quadratic term means change in log odds is positive for mean_age_cent below
-5.285, meaning mean_age between 0 and `r round(mean(merged_data$mean_age),3) - 5.285`.
In this interval, odds of expedition success is expected to multiply by a factor 
greater than 1 for a one year mean_age_cent increase. Change in log odds is negative
for mean_age_cent above -5.285, or mean age above `r round(mean(merged_data$mean_age),3)
- 5.285`, so odds of expedition success is expected to multiply by a factor less
than 1 for a one year mean_age_cent increase.

Hired_staff_deaths: For each additional death of hired staff, we would
expect, on average, that the odds of a successful expedition are
multiplied by `r round(exp(-0.81198), 3)`, holding all else constant.

Year: For each additional year since 1971, we would expect, on average,
that the odds of a successful expedition are multiplied by
`r round(exp(-0.00155), 3)`, holding all else constant.

Oxygen_usedTRUE: If oxygen were used, we would expect, on average, that
the odds of a successful expedition are multiplied by
`r round(exp(-0.77096), 3)`, holding all else constant.

height_metres:oxygen_usedTRUE: If oxygen were used, we would expect, on
average, that for every one meter increase in the height of the peak,
the odds of a successful expedition are multiplied by
`r round(exp(0.00098), 3)`, holding all else constant.

year:oxygen_usedTRUE: If oxygen were used, we would expect, on average,
that for every additional year since 1971, the odds of a successful
expedition are multiplied by `r round(exp(0.06035), 3)`, holding all
else constant.

**Findings**

The AUC value of 0.747 indicates that our model has a good fit to the data and 
can predict expedition success outcomes with high sensitivity and low false 
positive rates. The key predictors of expedition success found are all reasonable.
Climbers too young or too old would have a lower probability of success, explaining
the quadratic relationship. Oxygen use may indicate higher elevation or health 
complications, negatively impactimpacting on log odds of success. The surprising
finding that log odds of success declines with year may reflect the growing access
to mountain climbing. More amateur climbers in the Himalayas translates into 
decreasing log odds of success.

## 4. Discussion and Conclusion

Through this research we learned that the most significant predictors of expedition success were peak height, member count, mean member age, staff deaths,  year, and oxygen use. Higher peaks, staff, deaths, particularly young or old age, and use of oxygen reduced log odds of success. Expeditions further in the past and with more members had higher log odds of success. Our model with these predictors minimized BIC and created the simplest model for prediction. All the predictors identified were statistically significant, and the year variable, which initially had a high p-value was analyzed with a drop in deviance test, confirming the variable’s significance in the model. As expected, interaction terms between oxygen_use and height indicate reduced oxygen at higher elevations. The interaction term with year may also suggest changing oxygen technology over time.

Limitations: 

We did not examine all interaction terms between our predictor variables. These interaction terms might be impacting the accuracy of the model. Independence may also be influenced by particularly popular peaks.
We remove all data entries with one or more NAs. These entries might influence the final model prediction.
We did not explore other kinds of models other than logistic (after discovering linear models are probably not the best choice), but we fail to consider other possible models such as exponential and quadratic.
We were unable to examine the distributions of all predictor variables to see their effects on the model
I.e. There were likely more data from recent years

In the future, if we have more time to work on this project extensively, we might look at multiple interactions between the predictor variables we have selected. We can also look for alternative treatments of NA values, as well as  comparing our model with other models. Determining if any independence issues are caused by popular peaks would be informative, as well.


## 5. Appendix

**5.1 Empirical Logit Plots**

```{r}
par(mfrow = c(2,2))
emplogitplot1(success_expedition ~ year, data = merged_data, ngroups = 10)
emplogitplot2(success_expedition ~ year + oxygen_used, data = merged_data, 
              ngroups = 10, putlegend = "bottomright")

#par(mfrow = c(1,2))
emplogitplot1(success_expedition ~ height_metres_cent, data = merged_data, ngroups = 4)
emplogitplot2(success_expedition ~ height_metres_cent + oxygen_used, data = merged_data, 
              ngroups = 4, putlegend = "bottomright")

#par(mfrow = c(1,2))
emplogitplot1(success_expedition ~ member_count_cent, data = merged_data, ngroups = 8) 
emplogitplot2(success_expedition ~ member_count_cent + oxygen_used, data = merged_data, 
              ngroups = 8, putlegend = "bottomright")

#par(mfrow = c(1,2))
emplogitplot1(success_expedition ~ mean_age_cent, data = merged_data, ngroups = 8) 
emplogitplot2(success_expedition ~ mean_age_cent + oxygen_used, data = merged_data, 
              ngroups = 8, putlegend = "bottomright")

emplogitplot1(success_expedition ~ hired_staff_deaths, data = merged_data, ngroups = "all") 
# emplogitplot2(success_expedition ~ hired_staff_deaths + oxygen_used, data = merged_data, putlegend = "bottomright")
```
